Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-02-04 manuscript/climate_sensitivity_draft.Rmd
Update climate_sensitivity_draft.Rmd
workign on results, discussion

**Nobby questions using "useful predictor" as a term, instead using more jargon I think**
In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth: 1966, 1977, and 1999 (Figs. 1, S2). We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3). Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp. [*Cite Shenandoah paper, if accepted*; @twery_effects_1991].

@@

Larger-diameter trees to showed greater reductions in growth during drought, although there was no significant effect during 1977 or 1999 individually (Tables 1, 4). The same held true for $ln[H]$ as a univariate predictor (Tables 1, 4). When combined with other predictor variables in our multivariate models, the top models usually included an effect of $ln[H]$, with negative coefficient (Tables 1, 5).

@@

Hydraulic traits, including $XP$, $PLA_{dry}$, and $\pi_{tlp}$, explained variation in drought responses, whereas $LMA$ and $WD$ never significantly associated with $Rt$ in the univariate models (Tables 1,4,5). $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). $PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients. $\pi_{tlp}$ had no significant effect in the univariate models, but coefficients were consistently negative (Table 4). Whereas ring-porous species had highest $Rt$ overall and in the 1966 and 1999 droughts, they had lower $Rt$ in 1977. Results were similar in the context of multivariate models (Table 5), and $PLA_{dry}$, and $\pi_{tlp}$ were each included in the top overall model and in the top model for two of the three individual droughts (Table 5).

*Combined effects of tree size and traits*

*Our findings indicate that the greater growth reductions in large trees during drought (lower $Rt$) is driven by height itself, as opposed to more drought-sensitive traits in larger trees (Table 5). Tthere was little meaningful variation in traits with height at the community level (**Fig. S?**). When $ln[H]$ and hydraulic traits were considered together in multivariate models, the effect of $ln[H]$ on $Rt$ was consistently negative (Table 5). *

*Responses across droughts* {**remove this section, but first see if there's anything that should be preserved.**}

In the majority of cases, response directions were consistent across droughts in both univariate and multivariate models (Tables 1,4,5). However, there were a few exceptions-most commonly in the categorical variables ($CP$ and $XP$) but also for $PLA_{dry}$ in one multivariate model for the 1999 drought (Tables 4, 5). These differences may be spurious as opposed to statistically meaningful. Among the univariate models, there was no instance where predictor variables significantly improved the models of two different droughts (dAIC $\ge$ 2), but with contrasting coefficients (Table 4). Among the multivariate models, $CP$ was not consistently in the top models for any drought (Table 5), and $PLA_{dry}$ only appeared with a positive coefficient in the weakest predictor model of the six total models for the 1999 drought (contrasting with a negative coefficient in the univariate model; Table 4). It is probable that in the case of ring porosity, ring-porous species had lower $Rt$ than diffuse- and semi-ring- porous species in the 1977 drought, contrasting with higher resistance in 1966 and 1999 (Tables 4,5). Note, however, that $XP$ was not a signficant predictor on its own for the 1977 drought.

 For instance, $ln[H]$ and $PLA_{dry}$ had much stronger negative effects in 1966 than in the other two years, $ln[TWI]$ had the strongest negative effect in 1977, and $XP$ (higher $Rt$ among ring-porous trees) was strongest in 1999 (Tables 4,5). **see comment 37**

@@

Our results reveal how tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with the highest drought impacts felt by the most exposed and the most suppressed trees. This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**). **see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and larger rooting volum provided no advantage in the drier microenvironments. **see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy. Drought sensitivity was also not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate **see comment 41**; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature when not coupled with other disturbances. **see comment 42,43** While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), **see comment 44** supporting the observation that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, along with other non-climate factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5). The moderate nature of these droughts is helpful for our analysis because it considers only trees that survived all of these droughts, and we lack information on the trees that were killed prior to 2008. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in revision*).  Thus, the droughts considered here are substantially weaker than others that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and-to some extent-traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016). This is important because while our results are consistent with findings from more extreme, multiannual droughts, these extremes have not occurred in the Eastern United States similar to the West, and therefore we do not know exactly how forests would respond under higher mortality conditions [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016]. Given this consistency with our results, however, we believe this is an indication of a more intensive response.

@@

@@

The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time. At other sites, elliot et al analyzed a number of species as well, but responses were condensed into ring structure.  Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses-with both appearing in the top overall model and two of the three top models for each drought (Table 5)-is consistent with studies demonstrating that these are physiolgically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015]. Importantly, this finding indicates that both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--KAT/NOBBY/LAWREN**).

**see comment 48; the first paragraph was condensed here, but unsure where this should be put now**
Our analysis of hydraulic traits focused on species-level comparisons and did not characterize the role of variation with height. Further characterization of leaf hydraulic traits in relation to height and crown exposure would be valuable for enhancing our understanding of the interactive effects of tree height and traits on drought responses.

Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydraulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5). Although there were not significant differences in $Rt$ among the drought years, $Rt$ tended to be somewhat lower in 1966 (Fig. 1, Table 4). Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019]. **see comment 49** In 1977, which was the weakest drought in terms of PDSI (Fig. S2, Table S3), ring-porous species had lower $Rt$, contrasting with their higher $Rt$ in the other two droughts and perhaps indicating that ring-porosity provides less of an advantage in terms of drought adaptation under less extreme droughts, consistent with @elliott_forest_2015. In the 1999 drought, which was the most intense in terms of growing season PDSI and the second-longest in duration, ring porosity was the most important predictor, with diffuse-porous species negatively impacted (Fig. S2). Thus, responses differed with the nature of each drought. Of course, site characteristics also define the nature of droughts, and comparisons of size and trait effects across sites--and across more droughts at a single site--would be of great value to elucidating the mechanisms through which drought characteristics interact with driver variables to shape tree growth responses.

As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been disproportionately impacted in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable. Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young, non-dense forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age. Pioneer species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) **see comment 50** have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g. @rodriguez-caton_long-term_2015]. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.

>>>>

**Nobby questions using "useful predictor" as a term, instead using more jargon I think**
In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth: 1966, 1977, and 1999 (Figs. 1, S2). We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3). Rather, the severity of growth reduction may be explained in part by defoliation from gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which strongly impacted *Quercus* spp. [*Anderson-Teixeira et al., in review*; @twery_effects_1991].

@@

Larger-diameter trees to showed greater reductions in growth during drought, although there was no significant effect during 1977 or 1999 individually (Tables 1, 4). The same held true for $ln[H]$ as a univariate predictor (Tables 1, 4). When combined with other predictor variables, including species' traits, in our multivariate models, the top models usually included an effect of $ln[H]$, with negative coefficient (Tables 1, 5).

@@

Hydraulic traits, including $XP$, $PLA_{dry}$, and $\pi_{tlp}$, explained variation in drought responses (Tables 1,4,5). In the *single-variable tests*, $LMA$ and $WD$ never significantly associated with $Rt$ (Table 4) and were excluded from multivariable models. In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). $PLA_{dry}$ was a strong predictor for the 1966 drought and all droughts combined, with consistently negative coefficients (Table 4). Similarly $PLA_{dry}$ was consistently included, with negative coefficient, in multivariable models for the three droughts combined and for the 1966 and 1977 droughts individually (Table 5). $\pi_{tlp}$ had no significant effect in the univariate models, but coefficients were consistently negative (Table 4). It was, however, included in the top multivariate model for all droughts combined and for the 1977 and 1999 droughts individually (Table 5). Xylem porosity was not signficant for all droughts combined and had contrasting effects in the individual droughts: whereas ring-porous species had higher $Rt$ than diffuse- and semi-ring- porous species in the 1966 and 1999 droughts, they had lower $Rt$ in 1977 (Tables 4,5).

@@

Tree size, microenvironment, **see comment 38** and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with the highest drought impacts felt by the most exposed and the most suppressed trees. This is consistent with observations of both the former (e.g., [@suarez_factors_2004]; [@scharnweber_confessions_2019]) and the latter (**REFS**). **see comment 39** There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and larger rooting volum provided no advantage in the drier microenvironments. **see comment 40** The lower drought resistance of larger trees was not driven by a trend of more drought-sensitive species dominating the canopy. Drought sensitivity was also not consistently linked to species' $LMA$, $WD$, or xylem architecture. However, drought resistance was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts, indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

The droughts considered here were of similar severity (Fig. 1b) and fairly moderate **see comment 41**; droughts of this magnitude have occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019). Therefore, we excpect that most species are adapted, and individual trees acclimatized, to survive droughts of this nature when not coupled with other disturbances. **see comment 42,43** While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), **see comment 44** supporting the observation that these droughts did not induce extreme stress on the entire forest. It is likely for this reason, along with other non-climate factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5). The moderate nature of these droughts is helpful for our analysis because it considers only trees that survived all of these droughts, and we lack information on the trees that were killed prior to 2008. These are likely to be relatively modest in number, and local forest monitoring data stretching back to the late 1980s confirms that the 1999 drought did not trigger major declines in tree abundance or biomass (*Anderson-Teixeira et al., in review*).  Thus, the droughts considered here are substantially weaker than others that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]), many of which have shaped our understanding about the role of tree size [@bennett_larger_2015; @stovall_tree_2019] and-to some extent-traits underlying drought response mechanisms (@greenwood_tree_2017; @anderegg_meta-analysis_2016). This is important because while our results are consistent with findings from more extreme, multiannual droughts, these extremes have not occurred in the Eastern United States similar to the West, and therefore we do not know exactly how forests would respond under higher mortality conditions [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016]. Given this consistency with our results, however, we believe this is an indication of a more intensive response.

@@

Height and dominant canopy position had stronger negative effects in this drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing height's propensity to be a stronger predictor of mortality [@stovall_tree_2019]. **see comment 49**

@@

The development of tree-ring chronologies for all dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] made it possible to compare historical drought responses across 12 species and their associated traits at a single site for the first time. {*At other sites, elliot et al analyzed a number of species as well, but responses were condensed into ring structure.*}  Concerted measurement of leaf hydraulic traits of emerging importance [@scoffoni_leaf_2014; @bartlett_correlations_2016; @medeiros_extensive_2019] allowed novel insights into the role of hydraulic traits in shaping drought response. The finding that $PLA_{dry}$ and $\pi_{tlp}$ can be useful for predicting drought responses-with both appearing in the top overall model and two of the three top models for each drought (Table 5)-is consistent with studies demonstrating that these are physiologically meaningful traits linked to species distribution along moisture gradients [@medeiros_extensive_2019; @simeone_coupled_2019; @marechaux_drought_2015] {**Rosas et al NP about Spanish tree distributions; Fletcher et al AJB about Ceanothus**}. Importantly, this finding indicates that both $PLA_{dry}$ and $\pi_{tlp}$, which can be measured relatively easily  [@bartlett_rapid_2012; @scoffoni_leaf_2014], hold promise for predicting drought growth responses across species. The importance of linking species' traits to drought responses increases with tree species diversity; whereas it is feasible to study drought responses for all dominant species in most boreal and temperate forests (e.g., this study), this becomes difficult to impossible for species that don't form annual rings, and for diverse tropical forests. Although progress is being made for the tropics [@schongart_dendroecological_2017], a full linkage hydraulic traits to drought responses would be invaluable for forecasting how little-known species and whole forests will respond to future droughts (**REFS?--Nate Swenson papers? there was a recent review in TREE**).

*Although the physiological mechanisms discussed above lead to generally consistent directions of growth responses to tree height and hydraulic traits across droughts, indicating the universality of the underlying mechanisms, the relative importance of the drivers varied widely across droughts, indicating an interaction with drought characteristics (Tables 4-5). *

As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. Large trees have been disproportionately impacted in forests around the world [@bennett_larger_2015; @stovall_tree_2019], and we show, at least at this site, that this is primarily driven by their height with some contributions from canopy position. The distinction is important because it suggests that height *per se* makes trees vulnerable, even if their crowns are somewhat protected by neighbors, whereas solitary trees or the dominant trees in young regrowth forests should be less vulnerable. Considering just height and crown position, this would suggest that mature forests would be more vulnerable to drought than young, non-dense forests with short trees; however, soil water availability may limit the young forests [@bretfeld_plant_2018], and species traits often shift as forests age. Early successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) **see comment 50** have a mix of traits conferring drought tolerance and resistance (Table 3), and further research on how hydraulic traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g. @rodriguez-caton_long-term_2015]. In the meantime, the results of this study advance our knowledge of the factors conferring drought vulnerability and resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
